On the power of conditional independence testing under model-X \support{Support information of the article.}

### 2022-11-21 publication/ejs-sample.tex
Started on publication

\documentclass[12pt]{article}

 \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
 \usepackage{amssymb,amsfonts,graphicx,amsthm,nicefrac,mathtools,bm, bbm}
 \usepackage[hidelinks]{hyperref}
 \usepackage[normalem]{ulem} %to strike the words
 \usepackage{mathrsfs}
 \usepackage{pdflscape}

% \usepackage[numbers]{natbib}
 % \usepackage[english]{babel}
% \usepackage{times}
%  \usepackage[OT1]{fontenc}  
% \usepackage{bbm,mdframed}
 \usepackage[dvipsnames]{xcolor}
% \usepackage{xspace}
 % \usepackage{rotating,multirow}
 \usepackage{graphics,tikz}

\usepackage[]{caption}

\usepackage[linesnumbered,ruled]{algorithm2e}
\let\oldnl\nl% Store \nl in \oldnl
\newcommand{\nonl}{\renewcommand{\nl}{\let\nl\oldnl}}% Remove line number for one line
\makeatother

\usepackage[uniquename=false, uniquelist = false, url = false, doi = false, isbn = false, sorting = none, natbib = true, backend = bibtex, maxbibnames=10]{biblatex}

\addbibresource{~/.references/crtpower}
% \addbibresource{crtpower}

% \usepackage{grffile}
 % \usepackage{subcaption}
 % \usepackage{epstopdf} % for postscript graphics files

% \arxiv{arXiv:1803.06790}

 % \startlocaldefs

 %\setlength{\parindent}{10pt}
 %\setlength{\parskip}{5pt}

 %----------------------------------------------------------
 % Theorem etc macros
 %----------------------------------------------------------
 \newtheorem{theorem}{Theorem}
 \newtheorem{proposition}{Proposition}
 \newtheorem{claim}{Claim}
 \newtheorem{lemma}{Lemma}
 \newtheorem{corollary}{Corollary}
 \newtheorem{question}{Question}
 \newtheorem{assumption}{Assumption}
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}
   \newtheorem{setting}{Setting}
 \theoremstyle{remark}
 \newtheorem{remark}{Remark}
 \newtheorem{example}{Example}
 % \makeatletter

 %----------------------------------------------------------
 % Reference macros
 %----------------------------------------------------------
 \newcommand{\rmkref}[1]{Remark~\ref{rmk:#1}}
 \newcommand{\assumpref}[1]{Assumption~\ref{assump:#1}}
 \newcommand{\assumpsref}[1]{Assumptions~\ref{assump:#1}}
 \newcommand{\assumpssref}[1]{\ref{assump:#1}}
 \newcommand{\figref}[1]{Figure~\ref{fig:#1}}
 \newcommand{\figsref}[1]{Figures~\ref{fig:#1}}
 \newcommand{\figssref}[1]{\ref{fig:#1}}
 \newcommand{\chapref}[1]{Chapter~\ref{chap:#1}}
 \newcommand{\secref}[1]{Section~\ref{sec:#1}}
 \newcommand{\appref}[1]{Appendix~\ref{app:#1}}
 \newcommand{\eqssref}[1]{\ref{eqn:#1}}
 \newcommand{\defref}[1]{Definition~\ref{def:#1}}
 \newcommand{\exref}[1]{Example~\ref{ex:#1}}
 \newcommand{\clmref}[1]{Claim~\ref{clm:#1}}
 \newcommand{\lemref}[1]{Lemma~\ref{lem:#1}}
 \newcommand{\lemsref}[1]{Lemmas~\ref{lem:#1}}
 \newcommand{\lemssref}[1]{\ref{lem:#1}}
 \newcommand{\propref}[1]{Proposition~\ref{prop:#1}}
 \newcommand{\propsref}[1]{Propositions~\ref{prop:#1}}
 \newcommand{\propssref}[1]{\ref{prop:#1}}
 \newcommand{\thmref}[1]{Theorem~\ref{thm:#1}}
 \newcommand{\thmwref}[1]{Theorem~#1}
 \newcommand{\thmsref}[1]{Theorems~\ref{thm:#1}}
 \newcommand{\thmssref}[1]{\ref{thm:#1}}
% \newcommand{\corref}[1]{Corollary~\ref{cor:#1}}
 \newcommand{\tabref}[1]{Table~\ref{tab:#1}}
 \newcommand{\tabsref}[1]{Tables~\ref{tab:#1}}
 \newcommand{\tabssref}[1]{\ref{tab:#1}}
 \newcommand{\pwref}[1]{p.~{#1}}
 \newcommand{\ppwref}[1]{pp.~{#1}}
 \newcommand{\eqnref}[1]{\eqref{eqn:#1}}

 \setlength{\headheight}{35pt}

 %----------------------------------------------------------
 % Other macros
 %----------------------------------------------------------
 \newcommand{\iid}[0]{i.i.d.\xspace}
 \newcommand{\dist}[2]{\mathrm{dist}\!\left({#1},{#2}\right)} % distance
 \newcommand{\one}[1]{{\mathbbm{1}}_{{#1}}}
 \newcommand{\inner}[2]{\langle{#1},{#2}\rangle} % Inner product
 \newcommand{\norm}[1]{\left\lVert{#1}\right\rVert}
 %\newcommand{\PP}[1]{\mathbb{P}\left\{{#1}\right\}} % Probability
 \newcommand{\PP}[1]{\textnormal{Pr}\!\left\{{#1}\right\}} % Probability
 %\newcommand{\Pp}[2]{\mathbb{P}_{#1}\left\{{#2}\right\}} % Probability
 \newcommand{\EE}[1]{\mathbb{E}\left[{#1}\right]} % Expectation
 \newcommand{\Ep}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
 \newcommand{\EEst}[2]{\mathbb{E}\left[{#1}\ \middle| \ {#2}\right]} % Conditional expectation
 %\newcommand{\PPst}[2]{\mathbb{P}\left\{{#1}\ \middle| \ {#2}\right\}} % Conditional probability
 \newcommand{\PPst}[2]{\text{Pr}\!\left\{{#1}\ \middle| \ {#2}\right\}} % Conditional probability
 % \newcommand{\TV}[1]{\var\left({#1}\right)} % Variance
 % \newcommand{\Tp}[2]{\mathbb{V}_{#1}\left[{#2}\right]}
 % \newcommand{\TVst}[2]{\var\left({#1}\ \middle|\ {#2}\right)}
 % \renewcommand{\O}[1]{\mathcal{O}\left({#1}\right)}
 % \def\R{\mathbb{R}}
 % \def\Z{\mathbb{Z}}
 \newcommand{\ee}[1]{\mathbf{e}_{{#1}}}
 \newcommand{\ident}{\mathbf{I}}
 \newcommand{\ones}{\mathbf{1}}
 \newcommand{\zeros}{\mathbf{0}}
 \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
 \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
 \newcommand{\iidsim}{\stackrel{\mathrm{iid}}{\sim}}
 \newcommand{\indsim}{\stackrel{\independent}{\sim}}
 \newcommand{\notate}[1]{\textcolor{red}{\textbf{[#1]}}}
 \newcommand{\ignore}[1]{}
 \renewcommand{\Pr}[2]{\mathcal{P}_{{#1}}\left({#2}\right)} % projection
 \newcommand{\Prp}[2]{\mathcal{P}_{{#1}}^{\perp}\left({#2}\right)}
 \newcommand{\pr}[1]{\mathcal{P}_{{#1}}}
 \newcommand{\prp}[1]{\mathcal{P}_{{#1}}^{\perp}}
 \newcommand{\eps}{\epsilon}
 \newcommand{\lam}{\lambda}
 % \newcommand{\T}{\mathcal{T}}
 % \newcommand{\diff}[1]{\text{d}{#1}} % differential
 \let\emptyset\varnothing
 % \newcommand{\tf}{\widetilde{f}}
 % \newcommand{\tbeta}{\widetilde\beta}
 % \newcommand{\bhy}{\textnormal{BHY}}

@@

 % \usepackage[mmddyy]{datetime}
 % \renewcommand{\dateseparator}{.}

 %----------------------------------------------------------
 % Document-specific macros & settings
 %----------------------------------------------------------
 \newcommand{\nulls}{\mathcal{H}^0}
 \newcommand{\fdp}{\textnormal{FDP}}
 \newcommand{\fdr}{\textnormal{FDR}}
 \newcommand{\fdx}{\textnormal{FDX}}
 \newcommand{\fdphat}{\widehat{\textnormal{FDP}}}
 \newcommand{\fdpbar}{\overline{\textnormal{FDP}}}
 \newcommand{\fdphmu}{\widehat{\fdp^m_u}}
 \newcommand{\bh}{\textnormal{BH}}
 \newcommand{\pih}{\widehat{\pi}}
 \newcommand{\alphah}{\widehat{\alpha}}
 \newcommand{\vh}{\widehat{V}}
 %\def\L{\mathcal{L}}
 \def\S{\mathcal{S}}
 \def\F{\mathcal{F}}
 \def\G{\mathcal{G}}
 \def\M{\mathcal{M}}
 \def\H{\mathcal{H}}
 \def\cR{\mathcal{R}}
 \def\False{\mathcal{V}}

% \endlocaldefs

@@

\title{On the power of conditional independence testing \\ under model-X}
\author{Eugene Katsevich$^1$ and Aaditya Ramdas$^{1,2}$\\
	[10pt]
	\texttt{ekatsevi@wharton.upenn.edu},
	\texttt{aramdas@stat.cmu.edu} \\
	Department of Statistics and Data Science, University of Pennsylvania$^1$\\
	Department of Statistics and Data Science, 	Carnegie Mellon University$^2$\\
	Machine Learning Department, Carnegie Mellon University$^2$
}

\maketitle
\thispagestyle{empty}

\begin{abstract}
For testing conditional independence (CI) of a response $Y$ and a predictor $X$ given covariates $Z$, the recently introduced model-X (MX) framework has been the subject of active methodological research, especially in the context of MX knockoffs and their successful application to genome-wide association studies. In this paper, we study the power of MX CI tests, \textcolor{blue}{yielding quantitative insights into the role of machine learning and providing evidence in favor of using likelihood-based statistics in practice. Focusing on the conditional randomization test (CRT), we find that its conditional mode of inference allows us to reformulate it as testing a point null hypothesis involving the conditional distribution of $X$. The Neyman-Pearson lemma then implies that a likelihood-based statistic yields the most powerful CRT against a point alternative.} We also obtain a related optimality result for MX knockoffs. Switching to an asymptotic framework with arbitrarily growing covariate dimension, we derive an expression for the limiting power of the CRT against local semiparametric alternatives in terms of the prediction error of the machine learning algorithm on which its test statistic is based. Finally, we exhibit a resampling-free test with uniform asymptotic Type-I error control under the assumption that \textit{only the first two moments of $X$ given $Z$ are known}, a significant relaxation of the MX assumption. 
\end{abstract}

%	\begin{keyword}
%		\kwd{conditional independence testing}
%		\kwd{conditional randomization test}
%		\kwd{model-X}
%		\kwd{knockoffs}
%		\kwd{most powerful tests}
%		\kwd{local alternatives}
%		\kwd{dose response curve}
%	\end{keyword}

%\newpage 	
%\footnotesize
%\tableofcontents
%\normalsize

\color{black}

@@

%To keep an example in mind throughout the paper, consider a genome-wide association study (GWAS) based on $n$ individuals. For each individual, a disease status $\pry \in \{0,1\}$ is recorded along with the genotypes $\prx \in \{0,1,2\}$ at a single nucleotide polymorphism (SNP) of interest and $\prz \in \{0,1,2\}^p$ at all other SNPs. SNPs usually have only two possible alleles in the population (the \textit{minor allele} defined as the less common one), so the genotype at a SNP is defined as the number of minor alleles across maternal and paternal chromosomes. For example, $\prx = 1$ for an individual whose maternal and paternal chromosomes have a minor and a major allele, respectively, at the SNP of interest.

%
%consider a neuroscience experiment where we wish to determine if a particular \textit{voxel} (volume pixel) in the visual cortex responds preferentially to faces using an fMRI (functional magnetic resonance imaging) experiment involving showing a subject a sequence of $n$ images. Here, $\pry \in \mathbb R$ would be the fMRI brain recording 6 seconds after showing the image, $\prx \in \mathbb R$ would be a count of the number of faces in the image, and $\prz \in \mathbb R^{100}$ would include other visual features (counts of non-face objects, edge-detectors, color/brightness).
% \textcolor{red}{To keep an example in mind throughout the paper, consider $\pry \in \mathbb R^1$ cholesterol level, $\prx \in \{0,1,2\}^{10}$ the genotypes of an individual at 10 nearby polymorphic sites, and $\prz \in \{0,1,2\}^{500,000}$ the genotypes of the individual at other polymorphic sites across the genome.}

As formalized by Shah and Peters \cite{Shah2018}, the problem \eqref{conditional-independence} is fundamentally impossible without assumptions on the distribution $\mathcal L(\prx,\pry, \prz)$, in which case no asymptotically uniformly valid test of this hypothesis can have nontrivial power against \emph{any} alternative. In special cases, the problem is more tractable, for example if $\prz$ has discrete support, or if we were willing to make (semi)parametric assumptions on the form of $\mathcal L(\pry|\prx, \prz)$ (henceforth ``model-$Y|X$''). We will not be making such assumptions in this work.
% A complementary recent line of work has focused on shifting assumptions to the distribution $f_{\prx, \prz}$. In particular, 
Instead, we follow the lead of Candes et al. \cite{CetL16}, who proposed to avoid assumptions on $\mathcal L(\pry|\prx, \prz)$, but assume that we have access to $\mathcal L(\prx|\prz)$:\footnote{Candes et al. actually require that the full joint distribution $\mathcal L(\prx,\prz)$ is known, but this is because they also test for conditional associations between $\prz$ and $\pry$. We focus only on the relationship between $\prx$ and $\pry$ given $\prz$ and therefore require a weaker assumption.} 
\begin{equation} \label{eq:modelX}
\textit{model-X (MX) assumption}: \mathcal L(\prx| \prz)=  f^*_{\prx|\prz} \text{ for some known } f^*_{\prx|\prz}.\footnote{\textcolor{blue}{We implicitly assume that $\mathcal L$ has a density with respect to some dominating measure on $\mathbb R^{1+1+p}$, and that all conditional densities are well-defined almost surely. Here and throughout the paper, we identify probability distributions with their densities with respect to the appropriate dominating measure.}} 
\end{equation}
Candes et al.\ argue that while both model-$Y|X$ and MX are strong assumptions---especially when $p,d$ are large---in certain cases much more is known about $\prx | \prz$ than about $\pry|\prx, \prz$. In the aforementioned GWAS example, $\prx|\prz$ reflects the joint distribution of genotypes at SNPs across the genome, which is well described by hidden Markov models from population genetics \cite{SetC17}. On the other hand, the distribution $\pry|\prx,\prz$ represents the genetic basis of a complex trait, about which much less is known. In the context of (stratified) randomized experiments, the distribution $\mathcal L(\prx | \prz)$ is the propensity function \cite{Imai2004} (the analog of the propensity score for non-binary treatments \cite{Rosenbaum1983}) and is experimentally controlled. In general causal inference contexts, the MX assumption can be viewed as the assumption that the propensity function is known.

@@

\begin{equation}
	\label{eq:randomized-CRT}
	\phi^{\CRT}_T(\srx,\sry,\srz) \equiv 
	\begin{cases}
		1, \quad &\text{if }  T(\srx, \sry, \srz) > C(\sry,\srz); \\
		\gamma, \quad &\text{if }  T(\srx, \sry, \srz) = C(\sry,\srz); \\
		0, \quad &\text{if }T(\srx, \sry, \srz) < C(\sry,\srz).
	\end{cases}
\end{equation}
In order to accommodate discreteness, the CRT makes a randomized decision $\gamma$ when $T(\srx, \sry, \srz) = C(\sry,\srz)$ so that the size of the test is exactly $\alpha$.
%rejects with probability \[\gamma \equiv \frac{\alpha - \mathbb {P}[T(\srxk, \sry, \srz) > C(\sry,\srz)|\sry,\srz]}{\mathbb {P}[T(\srxk, \sry, \srz) = C(\sry,\srz)|\sry,\srz]}\] in cases when the test statistic is exactly equal to the rejection threshold. Note that the above probabilities in the definition of $\gamma$ are over the resampling distribution $\srxk\ |\ \sry, \srz$. 
In practice, the threshold $C(\sry,\srz)$ is approximated by computing $T(\srxk^b, \sry, \srz)$ for a large number $B$ of Monte Carlo resamples $\srxk^b \sim \srx|\srz$. % For $b = 1,\dots, B$, a resample $\srxk^b$ is drawn and the statistic $T(\srxk^b, \sry, \srz)$ is recomputed. Then, the CRT threshold is obtained through the empirical quantile
%\begin{equation}
%	\widehat C(\sry, \srz) \equiv Q_{1-\alpha}\left\{T(\srx, \sry, \srz), \{T(\srxk^b, \sry, \srz)\}_{b = 1, \dots, B}\right\}.
%	\label{empirical-quantile}
%\end{equation}
%If desired, a size $\alpha$ test can be obtained by computing $\gamma$ in the spirit of definition~\eqref{empirical-quantile}, but usually the CRT is implemented as a non-randomized test rejecting when $T(\srx, \sry, \srz) > \widehat C(\sry, \srz)$. 
%The randomization $\gamma$ is not applied in practice, the test rejecting when $T(\srx, \sry, \srz) > \widehat C(\sry, \srz)$. The finite-sample validity of the CRT for any $B \leq \infty$ and any $n,d,r,p$ follows from the fact that the resampled triples $(\srxk, \sry, \srz)$ are exchangeable with the original $(\srx, \sry, \srz)$ under the null, much like the argument used for a permutation test. 
For the sake of clarity, this paper considers only the ``infinite-$B$" version of the CRT as defined by~\eqref{upper-quantile} and \eqref{eq:randomized-CRT}. In the causal inference setting, the CRT can be viewed as a variant of Fisher's exact test for randomized experiments that incorporates strata of covariates \cite{Zheng2008,Hennessy2016}, basing inference on rerandomizing the treatment to the units.

@@

%The machine learning procedure used to fit $\widehat f_{\pry|\prx,\prz}$ may be time-consuming, and running the CRT usually requires refitting this model for each resample $(\srxk^b, \sry, \srz)$. This can make the CRT, e.g. as originally a computationally proposed \cite{CetL16}, an expensive procedure. However, this computational cost can be dramatically reduced using special kinds of test statistics $T$. For example, the HRT~\cite{Tansey2018} and the related digital twin test~\cite{Bates2020} proceed by training $\widehat f_{\pry|\prx,\prz}$ on an independent sample, and then defining
%\begin{equation}
%T^{\text{HRT}}(\srx, \sry, \srz) \equiv \sum_{i = 1}^n \log \widehat f_{\pry|\prx,\prz}(\sry_i|\srx_i,\srz_i),
%\label{HRT-statistic}
%\end{equation}
%i.e. the log-likelihood of the data under the trained model. Note that the HRT requires only one model fit and is therefore much faster than the originally-proposed CRT. Notably, the sample used for training $\widehat f_{\pry|\prx,\prz}$ need not come from the same distribution as $\mathcal L$ and can therefore be much larger. For example, in the context of the digital twin test \cite{Bates2020} it was pointed out that large observational datasets can be used to train $\widehat f_{\pry|\prx,\prz}$, while applying a variant of the CRT to a smaller experimental dataset to obtain a causal guarantee. Recently, the dCRT \cite{Liu2020} was proposed as an alternative test statistic for the CRT that allows in-sample training of $\widehat f_{\pry|\prx,\prz}$ while circumventing refitting.

%While there are now a variety of MX methodologies, they are unified by the following guiding principles:
%\begin{enumerate}
%	\item[P1.] (Machine learning) An approximation to the distribution $\pry|\prx,\prz$ is learned, and used to define a test statistic.
%	\item[P2.] (MX calibration) The test statistic is calibrated using the known $\prx|\prz$.
%	\item[P3.] (Conditional inference) Inference is valid conditionally on the observed $\sry$ and $\srz$. %\textcolor{red}{Recall (2)...}
%\end{enumerate}
%These are well-known to anyone familiar with MX methodology, but we find it useful to state them explicitly. P1 has empirically been found to be important for the power of MX methods: the accuracy of the machine learning method employed translates into improved power. Importantly, however, Type-I error control is guaranteed regardless of the quality of the learned distribution for $\pry|\prx,\prz$. This is because of P2: the calibration of the test statistic uses only the known distribution $\prx|\prz$. This leads to P3, conditional inference. Since only the distribution of $\srx$ is used for inference, MX methods calibrate their test statistics while holding $\sry$ and $\srz$ fixed. 
%
%Thus, several elegant methods have been designed, and important scientific applications have been identified where the MX assumption is reasonable. However, the search for powerful MX methodology has thus far not been grounded in a firm theoretical foundation. In particular, the following questions remain open:
%\begin{enumerate}
%	\item[Q1.] Are there ``optimal" test statistics for MX methods, in any sense?
%	\item[Q2.] To what extent can the MX assumption be weakened?
%	\item[Q3.] What is the precise connection between the performance of the machine learning (ML) step and the power of the resulting MX method?
%	\item[Q4.] Can the MX framework be used for estimation? If so, what estimands are of interest?
%\end{enumerate}
%Furthermore, this rapidly growing area remains somewhat disconnected from the vast theory on statistical testing and estimation. Can we leverage existing statistical theory to better understand MX methods? In this paper, we begin to address these questions. We summarize our main findings next.

@@

% Huang and Jansonrecently showed that finite-sample type-I error control is possible under only the assumption that the model for $(\prx, \prz)$ belongs to a known parametric family. We derive a complementary nonparametric result in Section~\ref{sec:weakening}, showing that if asymptotic validity suffices, then we need only the

% We show that the CRT, paired with the \textit{generalized covariance measure} statistic of Shah and Peters \cite{Shah2018}, retains asymptotic Type-I error control under the MX(2) assumption. Requiring knowledge of just the first two moments of the conditional distribution $\prx|\prz$, rather than the distribution itself, promises to broaden the scope of application of MX-style methodology.

%This result clearly shows how the prediction error of $\widehat g$ impacts the asymptotic efficiency of the CRT. Unlike parallel work in causal inference, the above power result holds conditionally on $\sry, \srz$, almost surely.

%\paragraph{A4: Nonparametric targets can be consistently estimated under MX using causal inference tools.}
%
%All MX methodology thus far has focused on testing. In Section~\ref{sec:causal}, we approach the subject of estimation under MX by drawing close connections between MX and causal inference, since estimation is a major thrust of the latter field. The MX setting is indeed remarkably similar to that of a randomized experiment with continuous or multivariate treatment and covariates, where the \textit{propensity function} (a generalization of the propensity score) is known. The MX calibration principle (P2) is directly analogous to using randomization as the basis for inference, in both cases obviating the need for any assumptions on $\pry|\prx,\prz$. The question of even defining the target of estimation is a nontrivial one, and several options are possible. A nonparametric generalization of the parameter $\beta$ in the model~\eqref{parametric-alternative-intro} is
%%\begin{equation*}
%$
%\beta(\mathcal L) ~ \equiv ~
%\mathbb E_{\mathcal L}[\text{Var}[\prx|\prz]]^{-1}
%\mathbb E_{\mathcal L}[\text{Cov}[\prx,\pry|\prz]],
%$
%%\end{equation*}
%which is related to the \textit{variance-weighted average treatment effect} in causal inference \cite{Li2011}. 
%%We complement existing results on estimation of this quantity (e.g., \cite{Robins2009, Li2011}) by showing that a standard regression-based estimator is consistent, conditionally on $\sry,\srz$, almost surely in $\sry,\srz$. 
%We also suggest that the \textit{dose-response function} \cite{Hirano2004} may be an interesting estimand in the MX framework; this quantity can be estimated using causal inference tools \cite{Kennedy2017} if the propensity function is known.

@@

%\section{Conditional randomization test methodology} \label{sec:background}
%
%\subsection{Holdout Randomization Test (HRT)} \label{sec:HRT}
%
%The HRT first randomly splits the data into training and test sets of possibly unequal sizes. The model $\widehat \theta = \widehat \theta_{\text{train}}$ is fit on the training set, and then the CRT is applied on the test set, conditionally on $\widehat \theta_{\text{train}}$. In this paper, we keep the training set implicit and use $(\srx, \sry, \srz)$ for the test set. In this notation, the authors \cite{Tansey2018} proposed the following test statistic:
%\begin{equation}
%T(\srx, \sry, \srz) \equiv \sum_{i = 1}^n \log f_{\pry|\prx,\prz}^{\widehat \theta_{\text{train}}}(\sry_i|\srx_i,\srz_i),
%\label{HRT-statistic}
%\end{equation}
%i.e. the log-likelihood of the data under the trained model. Note that the HRT requires only one model fit and is therefore much faster than the full CRT. In fact, the HRT is simply a special case of the CRT based on a test statistic that is easy to compute. The HRT trades some statistical efficiency for this computational speed due to its use of sample splitting. We note that the recently proposed digital twin test for causal inference from trio studies \cite{Bates2020} is an instance of the HRT, since the multivariate model employed there is fit once on a separate source of data from the CRT resampling.
%
%
%\subsection{The Hybrid and MX(2) CRTs \textcolor{red}{do we need this business?}} \label{sec:hybrid-crt}
%The recently-proposed distilled CRT methodology \cite{Liu2020}  njoys computational speed and power approaching those of the HRT and CRT, respectively. This method is based on the observation that, if a part of $\widehat \theta$ is fit on only $(\sry, \srz)$, then there is no need to refit it upon resampling. Therefore, the in-sample data can be used efficiently as long as the computationally intensive part excludes $\srx$. Leveraging out-of-sample data as well, we obtain the Hybrid CRT (Algorithm~\ref{alg:hybrid-crt}), which we formulate in terms of the semiparametric model~\eqref{parametric-alternative-intro}. It is a hybrid of the HRT and CRT, as it fits $g$ only once (like the HRT) but refits $\beta$ after each resampling of $\srx_i | \srz_i$ (like the CRT). In contrast, the HRT fits $\beta,g$ just once, while the full CRT refits $\beta,g$ together in every resampling step.
%
%\noindent
%\begin{center}
%	\begin{minipage}{\linewidth}
%		\begin{algorithm}[H]
%			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
%			\KwData{
%				$\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, loss function $\ell$, machine learning method $g$
%			}
%			Obtain $\widehat g$ by fitting $g$ on $\{(\sry_i,\srz_i)\}_{i=1}^n$ or a different dataset or both\;

>>>>

\documentclass[ejs]{imsart}

\RequirePackage{amsthm,amsmath,amsfonts}
\RequirePackage[authoryear]{natbib}
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\RequirePackage{graphicx}
\RequirePackage{bm, bbm}

% settings
\pubyear{2022}
\volume{0}
\issue{0}
\firstpage{1}
\lastpage{8}
\arxiv{2005.05506}

\startlocaldefs
\numberwithin{equation}{section}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
 \newtheorem{proposition}{Proposition}
 \newtheorem{claim}{Claim}
 \newtheorem{lemma}{Lemma}
 \newtheorem{corollary}{Corollary}
 \newtheorem{question}{Question}
 \newtheorem{assumption}{Assumption}
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}
   \newtheorem{setting}{Setting}
 \theoremstyle{remark}
 \newtheorem{remark}{Remark}
 \newtheorem{example}{Example}

\newcommand{\iid}[0]{i.i.d.\xspace}
 \newcommand{\dist}[2]{\mathrm{dist}\!\left({#1},{#2}\right)} % distance
 \newcommand{\one}[1]{{\mathbbm{1}}_{{#1}}}
 \newcommand{\inner}[2]{\langle{#1},{#2}\rangle} % Inner product
 \newcommand{\norm}[1]{\left\lVert{#1}\right\rVert}
 \newcommand{\PP}[1]{\textnormal{Pr}\!\left\{{#1}\right\}} % Probability
 \newcommand{\EE}[1]{\mathbb{E}\left[{#1}\right]} % Expectation
 \newcommand{\Ep}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
 \newcommand{\EEst}[2]{\mathbb{E}\left[{#1}\ \middle| \ {#2}\right]} % Conditional expectation
 \newcommand{\PPst}[2]{\text{Pr}\!\left\{{#1}\ \middle| \ {#2}\right\}} % Conditional probability
 \newcommand{\ee}[1]{\mathbf{e}_{{#1}}}
 \newcommand{\ident}{\mathbf{I}}
 \newcommand{\ones}{\mathbf{1}}
 \newcommand{\zeros}{\mathbf{0}}
 \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
 \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
 \newcommand{\iidsim}{\stackrel{\mathrm{iid}}{\sim}}
 \newcommand{\indsim}{\stackrel{\independent}{\sim}}
 \newcommand{\notate}[1]{\textcolor{red}{\textbf{[#1]}}}
 \newcommand{\ignore}[1]{}
 \renewcommand{\Pr}[2]{\mathcal{P}_{{#1}}\left({#2}\right)} % projection
 \newcommand{\Prp}[2]{\mathcal{P}_{{#1}}^{\perp}\left({#2}\right)}
 \newcommand{\pr}[1]{\mathcal{P}_{{#1}}}
 \newcommand{\prp}[1]{\mathcal{P}_{{#1}}^{\perp}}
 \newcommand{\eps}{\epsilon}
 \newcommand{\lam}{\lambda}

 \let\emptyset\varnothing

@@

\endlocaldefs

@@

\begin{frontmatter}
\title{On the power of conditional independence testing under model-X \support{Support information of the article.}}
\runtitle{A Sample Document}

\begin{aug}
\author{\fnms{Eugene} \snm{Katsevich}\thanksref{t1,t2}\ead[label=e1]{ekatsevi@wharton.upenn.edu}}
\and
\author{\fnms{Aaditya} \snm{Ramdas}\thanksref{t3}\ead[label=e2]{aramdas@stat.cmu.edu}}

\address{Department of Statistics and Data Science, University of Pennsylvania\\
Department of Statistics and Data Science, Carnegie Mellon University\\
Machine Learning Department, Carnegie Mellon University \\
\printead{e1,e2}}

\thankstext{t1}{Some comment}
\thankstext{t2}{First supporter of the project}
\thankstext{t3}{Second supporter of the project}
\runauthor{F. Author et al.}

\end{aug}

\begin{abstract}
       For testing conditional independence (CI) of a response $Y$ and a predictor $X$ given covariates $Z$, the model-X (MX) framework has been the subject of active methodological research, especially in the context of MX knockoffs and their application to genome-wide association studies. In this paper, we study the power of MX CI tests, yielding quantitative insights into the role of machine learning and providing evidence in favor of using likelihood-based statistics in practice. Focusing on the conditional randomization test (CRT), we find that its conditional mode of inference allows us to reformulate it as testing a point null hypothesis involving the conditional distribution of $X$. The Neyman-Pearson lemma implies that a likelihood-based statistic yields the most powerful CRT against a point alternative. We obtain a related optimality result for MX knockoffs. Switching to an asymptotic framework with arbitrarily growing covariate dimension, we derive an expression for the power of the CRT against local semiparametric alternatives in terms of the prediction error of the machine learning algorithm on which its test statistic is based. Finally, we exhibit a resampling-free test with uniform asymptotic Type-I error control under the assumption that only the first two moments of $X$ given $Z$ are known.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{60K35}
\kwd{60K35}
\kwd[; secondary ]{60K35}
\end{keyword}

\begin{keyword}
\kwd{sample}
\kwd{\LaTeXe}
\end{keyword}
\tableofcontents
\end{frontmatter}

@@

As formalized by Shah and Peters \cite{Shah2018}, the problem \eqref{conditional-independence} is fundamentally impossible without assumptions on the distribution $\mathcal L(\prx,\pry, \prz)$, in which case no asymptotically uniformly valid test of this hypothesis can have nontrivial power against \emph{any} alternative. In special cases, the problem is more tractable, for example if $\prz$ has discrete support, or if we were willing to make (semi)parametric assumptions on the form of $\mathcal L(\pry|\prx, \prz)$ (henceforth ``model-$Y|X$''). We will not be making such assumptions in this work. Instead, we follow the lead of Candes et al. \cite{CetL16}, who proposed to avoid assumptions on $\mathcal L(\pry|\prx, \prz)$, but assume that we have access to $\mathcal L(\prx|\prz)$:\footnote{Candes et al. actually require that the full joint distribution $\mathcal L(\prx,\prz)$ is known, but this is because they also test for conditional associations between $\prz$ and $\pry$. We focus only on the relationship between $\prx$ and $\pry$ given $\prz$ and therefore require a weaker assumption.} 
\begin{equation} \label{eq:modelX}
\textit{model-X (MX) assumption}: \mathcal L(\prx| \prz)=  f^*_{\prx|\prz} \text{ for some known } f^*_{\prx|\prz}.\footnote{\textcolor{blue}{We implicitly assume that $\mathcal L$ has a density with respect to some dominating measure on $\mathbb R^{1+1+p}$, and that all conditional densities are well-defined almost surely. Here and throughout the paper, we identify probability distributions with their densities with respect to the appropriate dominating measure.}} 
\end{equation}
Candes et al.\ argue that while both model-$Y|X$ and MX are strong assumptions---especially when $p,d$ are large---in certain cases much more is known about $\prx | \prz$ than about $\pry|\prx, \prz$. In the aforementioned GWAS example, $\prx|\prz$ reflects the joint distribution of genotypes at SNPs across the genome, which is well described by hidden Markov models from population genetics \cite{SetC17}. On the other hand, the distribution $\pry|\prx,\prz$ represents the genetic basis of a complex trait, about which much less is known. In the context of (stratified) randomized experiments, the distribution $\mathcal L(\prx | \prz)$ is the propensity function \cite{Imai2004} (the analog of the propensity score for non-binary treatments \cite{Rosenbaum1983}) and is experimentally controlled. In general causal inference contexts, the MX assumption can be viewed as the assumption that the propensity function is known.

@@

\begin{equation}
	\label{eq:randomized-CRT}
	\phi^{\CRT}_T(\srx,\sry,\srz) \equiv 
	\begin{cases}
		1, \quad &\text{if }  T(\srx, \sry, \srz) > C(\sry,\srz); \\
		\gamma, \quad &\text{if }  T(\srx, \sry, \srz) = C(\sry,\srz); \\
		0, \quad &\text{if }T(\srx, \sry, \srz) < C(\sry,\srz).
	\end{cases}
\end{equation}
In order to accommodate discreteness, the CRT makes a randomized decision $\gamma$ when $T(\srx, \sry, \srz) = C(\sry,\srz)$ so that the size of the test is exactly $\alpha$. In practice, the threshold $C(\sry,\srz)$ is approximated by computing $T(\srxk^b, \sry, \srz)$ for a large number $B$ of Monte Carlo resamples $\srxk^b \sim \srx|\srz$. For the sake of clarity, this paper considers only the ``infinite-$B$" version of the CRT as defined by~\eqref{upper-quantile} and \eqref{eq:randomized-CRT}. In the causal inference setting, the CRT can be viewed as a variant of Fisher's exact test for randomized experiments that incorporates strata of covariates \cite{Zheng2008,Hennessy2016}, basing inference on rerandomizing the treatment to the units.

@@

@@

@@
